---
title: "Germination fluidics"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---
Distribution of events following virospore germination

## Clear workspace and set directory

```{r setup}
rm(list = ls())
getwd()
setwd("~/GitHub/viro.fluidics/")
```

```{r}
## Load packages and functions
library("bbmle")
library("ggplot2")
require("png")
require("dplyr")
require("grid")
```

```{r}
# Data
```{r}
fluid <- read.csv("data/fluidics.csv", header = TRUE, sep = ",", na.strings = "N/A")
data <- data.frame(fluid$treat, fluid$germ.time)
colnames(data) <- c("treat", "germ")
# Remove rows with NA values
data <- na.omit(data)
```

```{r}
# Rescale the data for numerical stability
data$germ_scaled <- scale(data$germ)

# Extended GEV likelihood function with debugging
gev_likelihood_treatment_debug <- function(shape, location, scale, 
                                           treat_shape = 0, treat_location = 0, treat_scale = 0) {
  data$treat_effect <- ifelse(data$treat == "infect", 1, 0)
  
  # Compute treatment-adjusted parameters
  shape_eff <- shape + treat_shape * data$treat_effect
  location_eff <- location + treat_location * data$treat_effect
  scale_eff <- scale + treat_scale * data$treat_effect

  if (any(scale_eff <= 0)) return(Inf)
  max_germ <- max(data$germ_scaled, na.rm = TRUE)
  if (any(shape_eff < 0 & max_germ > location_eff + scale_eff / abs(shape_eff))) return(Inf)

  ll <- tryCatch({
    sum(mapply(
      function(x, loc, sc, sh) {
        if (sc <= 0) return(Inf)
        dgev(x, shape = sh, loc = loc, scale = sc, log = TRUE)
      },
      x = data$germ_scaled,
      loc = location_eff,
      sc = scale_eff,
      sh = shape_eff
    ), na.rm = TRUE)
  }, error = function(e) Inf)

  if (!is.finite(ll)) return(Inf)
  -ll
}
```


# Define parameter bounds
```{r}
base_lower <- list(shape = -2, location = -3, scale = 0.01)
base_upper <- list(shape = 2, location = 3, scale = 10)
treat_lower <- c(base_lower, treat_shape = -1, treat_location = -1, treat_scale = -1)
treat_upper <- c(base_upper, treat_shape = 1, treat_location = 1, treat_scale = 1)
```


# Fit models
```{r}
fit_null <- mle2(gev_likelihood_treatment_debug,
                 start = list(shape = 0.1, location = mean(data$germ_scaled), scale = sd(data$germ_scaled)),
                 method = "Nelder-Mead",
                 control = list(maxit = 2000))

fit_full <- mle2(gev_likelihood_treatment_debug,
                 start = list(shape = 0.1, location = mean(data$germ_scaled), scale = sd(data$germ_scaled),
                              treat_shape = 0, treat_location = 0, treat_scale = 0),
                 method = "Nelder-Mead",
                 control = list(maxit = 2000))
```

```{r}
# Extract fitted parameters from the best model
fitted_params <- coef(fit_full)

# Back-transform fitted parameters to the original scale
shape_infect <- fitted_params["shape"] + fitted_params["treat_shape"]
location_infect <- fitted_params["location"] + fitted_params["treat_location"]
scale_infect <- fitted_params["scale"] + fitted_params["treat_scale"]

shape_control <- fitted_params["shape"]
location_control <- fitted_params["location"]
scale_control <- fitted_params["scale"]

# Create data for GEV density lines
x_values <- seq(min(data$germ), max(data$germ), length.out = 500)
infect_density <- dgev((x_values - mean(data$germ)) / sd(data$germ), 
                       shape = shape_infect, loc = location_infect, scale = scale_infect) / sd(data$germ)
control_density <- dgev((x_values - mean(data$germ)) / sd(data$germ), 
                        shape = shape_control, loc = location_control, scale = scale_control) / sd(data$germ)
```

# Create a dataframe for the density lines
```{r}
density_data <- data.frame(
  x = rep(x_values, 2),
  density = c(infect_density, control_density),
  treat = rep(c("infected", "uninfected"), each = length(x_values))
)
```


# Plot histograms with GEV fits
```{r}

png(filename = "~/GitHub/viro.fluidics/plots/GVE.png",
    width = 1200, height = 1200, res = 96 * 2)

plot <- ggplot() +
  geom_histogram(data = subset(data, treat == "control"), aes(x = germ, y = ..density..), 
                 fill = "red", alpha = 0.5, bins = 30, color = "black") +  # Uninfected histogram
  geom_histogram(data = subset(data, treat == "infect"), aes(x = germ, y = ..density..), 
                 fill = "blue", alpha = 0.5, bins = 30, color = "black") +  # Infected histogram
  geom_line(data = density_data, aes(x = x, y = density, color = treat), 
            linewidth = 0.85, linetype = "solid") +  # Dashed lines added
  scale_color_manual(values = c("infected" = "blue", "uninfected" = "red"),
                     labels = c("infected", "uninfected")) +
  labs(x = "Germination time (min)", y = "Density") +
  theme_minimal() +
theme(
  legend.title = element_blank(),                    # Remove legend title
  legend.text = element_text(size = 14),             # Increase legend text size
  legend.position = c(0.5, 0.85),                    # Move legend to the top
  legend.justification = "center",                   # Center-align legend
  panel.grid = element_blank(),                      # Remove gridlines
  axis.line = element_line(color = "black", linewidth = 0.8), # Axis lines
  axis.ticks = element_line(color = "black", linewidth = 0.8), # Tick marks
  axis.ticks.length = unit(0.4, "cm"),               # Double tick mark length
  axis.title.x = element_text(size = 16, margin = margin(t = 25)), # Combine size and margin for x-axis title
  axis.title.y = element_text(size = 16, margin = margin(r = 25)), # Combine size and margin for y-axis title
  axis.text.x = element_text(size = 14),
  axis.text.y = element_text(size = 14),
  legend.key = element_rect(fill = NA, color = NA),  # Remove legend symbol borders
  plot.margin = margin(20, 20, 20, 20)               # Add more buffer space
)


# Print the plot to the PNG device
print(plot)

# Close the device
dev.off()

# Display the saved image in RStudio
img <- readPNG("~/GitHub/viro.fluidics/plots/GVE.png")
grid.raster(img)
```


